---
title: "Instrumental Variable Analysis with Piperacillin-Tazobactam Shortage"
author: "Fergus Hamilton"
format: html
embed-resources: true
execute:
  echo: false
  warning: false
---

## Introduction

In this document, we perform an instrumental variable (IV) analysis using a simulated dataset to illustrate why adjusting for Metronidazole (MTZ) is a bad idea in the context of a piperacillin-tazobactam (pip-taz) shortage when MTZ is commonly used instead. We will also provide a Directed Acyclic Graph (DAG) to visually explain the relationships between the variables.

## Directed Acyclic Graph (DAG)

Below is a DAG that represents the relationships between the variables in our analysis:

```{r}
library(dagitty)
library(ggdag)
library(ggplot2)

dag <- dagitty('dag {
  Z [exposure]
  X [outcome]
  M [latent]
  Y [outcome]
  U [latent]
  
  Z -> X
  Z -> M
  U -> Y
  U -> M

}')

ggdag(dag, text = TRUE, layout = "circle") +
  theme_dag() +
  ggtitle("Simple causal model")

```

In this DAG:

```         
•   Z is the instrument (Pip-Taz Shortage).
•   X is the exposure (use of cefipime)
•   M is the Metronidazole use.
•   Y is the outcome (e.g. mortality, or length of stay)
•   U is an unmeasured confounder.
```

Adjusting for M in this context is problematic because M is influenced by both Z and U. By adjusting for M, we can introduce bias due to conditioning on a collider, which opens up a backdoor path that was otherwise blocked.

```{r}

# Load necessary libraries
library(dplyr)
library(broom)
library(AER)  # For IV regression

# Set seed for reproducibility
set.seed(42)

# Simulation parameters
n <- 50000
p <- 0.5  # Probability of Z (Pip-Taz Shortage)
q <- 0.5  # Probability of U (Confounder)

# Generate data
Z <- rbinom(n, 1, p)  # Pip-Taz Shortage (Instrument)
U <- rbinom(n, 1, q)  # Confounder

# Generate X based on Z 
X <- rbinom(n, 1, plogis(0.5 + 3 * Z))

# Generate Metronidazole (M) based on Z and U with strong effects
M <- rbinom(n, 1, plogis(0.5 + 2 * Z + 2 * U))

# Generate survival probability (Y) based on U only with strong effects, and no causal effect of X
Y <- rbinom(n, 1, plogis(0.5 + 2 * U))

# Create DataFrame
data <- data.frame(Z = Z, U = U, X = X, M = M, Y = Y)

# IV regression without adjusting for M
iv_model1 <- ivreg(Y ~ X | Z, data = data)
summary1 <- summary(iv_model1)

# IV regression adjusting for M
iv_model2 <- ivreg(Y ~ X + M | Z + M, data = data)
summary2 <- summary(iv_model2)

# Display the results
print("IV Regression without adjusting for M")
print(summary1)

print("IV Regression adjusting for M")
print(summary2)



# Extract coefficients and confidence intervals
estimates <- data.frame(
  model = c("Without adjusting for MTZ", "Adjusting for MTZ"),
  estimate = c(coef(summary1)[2, 1], coef(summary2)[2, 1]),
  conf.low = c(confint(iv_model1)[2, 1], confint(iv_model2)[2, 1]),
  conf.high = c(confint(iv_model1)[2, 2], confint(iv_model2)[2, 2])
)

# Create forest plot
ggplot(estimates, aes(x = model, y = estimate)) +
  geom_point(size = 4) +
  geom_hline(aes(yintercept = 0), col = "red", lty = "dashed") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(title = "Estimates of the effect of pip-taz on any outcome",
       x = "Model",
       y = "Estimate (95% CI)") +
  theme_bw() +
  coord_flip()
```

Adjusting for Metronidazole (MTZ) in the analysis introduces bias because MTZ is affected by both the instrument (Z) and the confounder (U). When we condition on MTZ, we inadvertently open a backdoor path from Z to Y through U, which can introduce spurious associations and bias the estimate of the effect of X on Y.

To understand this, think of MTZ as a marker for more severe illness. Patients who are more severely ill are more likely to receive MTZ. In our DAG, MTZ (M) is also influenced by the instrument (Z, pip-taz shortage) because during a shortage, doctors prescribe MTZ more often. This is shown in the paper. MTZ is influenced by the confounder (U), which represents the severity of illness.

When we adjust for MTZ, we are essentially saying, “Let’s compare patients who received MTZ to those who didn’t, within each level of MTZ use.” However, since MTZ use is driven by both the shortage and the severity of illness, this comparison mixes up the effects of the shortage and the illness severity.

In simpler terms:

• Without adjusting for MTZ: We are comparing mortality rates between those who had used cefipime and those who didn’t, while ignoring MTZ use.

• Adjusting for MTZ: We are comparing mortality rates between those who had access to pip-taz and those who didn’t, within the groups of those who used MTZ and those who didn’t. This comparison is flawed because MTZ use itself is influenced by how sick patients are, which is not directly related to the pip-taz shortage.


Therefore, adjusting for MTZ breaks the assumption that the instrument (Z) only affects the outcome (Y) through the exposure (cefipime usage), leading to biased estimates.

Conclusion

Our analysis illustrates that adjusting for colliders such as MTZ can introduce bias in instrumental variable analyses. In the context of a piperacillin-tazobactam shortage where MTZ is commonly used as an alternative, it is crucial to avoid adjusting for MTZ to maintain the validity of the IV assumptions.
